We want to simplify a mesh while preserving its \emph{mechanical responsiveness},
formulating a quadric error analogous to the geometric QEM of Garland \& Heckbert.
This section builds the analogy from the ground up, identifies the natural
``plane quadric'' for the mechanical case, and is honest about where the geometric
and mechanical settings genuinely diverge.

\subsection{What the QEM plane quadric actually is}

Stripped to its skeleton, the geometric face quadric is a \textbf{squared linear residual}:
\begin{equation}
  E_f(\mathbf{x}) = r_f(\mathbf{x})^2, \qquad r_f(\mathbf{x}) = \mathbf{n}^T\mathbf{x} + d.
\end{equation}
\begin{itemize}
  \item The \textbf{variable} is an absolute point position $\mathbf{x}\in\mathbb{R}^3$.
  \item The \textbf{residual} $r_f$ is the signed distance to the plane---a \emph{linear
        functional} of the variable that is zero on the ``correct'' set (the plane itself).
  \item The matrix $\mathbf{n}\mathbf{n}^T$ is rank-1 and positive semi-definite; its
        \textbf{null space is the in-plane directions}---the motions that cost nothing.
  \item Accumulation ($\sum_f$) works because every face's residual is written in the
        \emph{same} variable $\mathbf{x}$.
\end{itemize}
So QEM follows a template: \emph{pick a quantity that should be preserved (flatness),
write its violation as a squared linear residual, sum the residuals.}

\subsection{Transplanting the template to mechanics}

The quantity we want to preserve is \emph{elastic response}, encoded by the element
energy. For the CST membrane element,
\begin{equation}
  W_e(\mathbf{d}) = \tfrac12\,\mathbf{d}^T K_e\,\mathbf{d}
                  = \tfrac12\,A\,t\,\boldsymbol\varepsilon^T D\,\boldsymbol\varepsilon,
  \qquad \boldsymbol\varepsilon = B\mathbf{d}.
\end{equation}
Because $W_e$ is quadratic in the displacement, $K_e$ is exactly its \textbf{Hessian}
$\partial^2 W_e/\partial\mathbf{d}^2$: the element stiffness \emph{is} the curvature of the
elastic energy, and assembly $K=\sum_e K_e$ is the Hessian of the total energy (constant,
because linear elasticity is quadratic; equivalently the tangent stiffness at the rest
state). Everything downstream is this Hessian, restricted to a subspace and tracked as the
mesh coarsens.
This is \emph{already} a squared linear residual, but with a non-trivial metric.
Factor the constitutive matrix as $D = L L^T$ (Cholesky, i.e.\ $D^{1/2}$) and define
$M = \sqrt{A t}\,L^T B \in \mathbb{R}^{3\times 9}$. Then
\begin{equation}
  W_e(\mathbf{d}) = \tfrac12\sum_{k=1}^{3}\big(\mathbf{m}_k^T\mathbf{d}\big)^2,
\end{equation}
where $\mathbf{m}_k^T$ are the rows of $M$. \textbf{Each term is a generalized plane
quadric.} The strain plays the role of distance, and the constitutive matrix $D$ is the
metric that geometric QEM leaves as the identity.

The bending hinge is even cleaner---it is \emph{literally} a single plane quadric:
\begin{equation}
  W_b(\mathbf{d}) = \tfrac12\,c\,(\mathbf{g}^T\mathbf{d})^2
                  = \tfrac12\,\big(\underbrace{\sqrt{c}\,\mathbf{g}}_{\text{``normal''}}{}^{\!T}\mathbf{d}\big)^2.
\end{equation}
The linearized dihedral angle $\mathbf{g}^T\mathbf{d}$ \emph{is} the signed-distance
residual; $\sqrt{c}\,\mathbf{g}$ \emph{is} the plane normal---only it lives in $12$-D
displacement space instead of $3$-D position space.

\subsection{The dictionary}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{0.46\linewidth}|p{0.46\linewidth}|}
\hline
\textbf{QEM (geometry)} & \textbf{Mechanical analog} \\
\hline
variable: point position $\mathbf{x}\in\mathbb{R}^3$
  & nodal displacement $\mathbf{d}\in\mathbb{R}^{9}$ (or $\mathbb{R}^{12}$ hinge) \\
\hline
residual: signed distance $\mathbf{n}^T\mathbf{x}+d$
  & strain $\boldsymbol\varepsilon = B\mathbf{d}$ (resp.\ $\mathbf{g}^T\mathbf{d}$) \\
\hline
metric on residual: identity
  & constitutive $D$ (scaled by $A t$) \\
\hline
face quadric: $\mathbf{n}\mathbf{n}^T$, rank $1$
  & element stiffness $K_e = A t\,B^T D B$, rank $3$ (hinge: rank $1$) \\
\hline
null space: in-plane slides (don't matter)
  & rigid-body modes (don't matter) \\
\hline
accumulation: $\sum_f$ on shared $\mathbf{x}$
  & assembly $K = \sum_e K_e$ on shared DOFs \\
\hline
\end{tabular}
\end{center}

\noindent\textbf{The mechanical ``plane quadric'' is the element stiffness $K_e$}, viewed
as a squared strain-residual in the energy metric $D$. It is positive semi-definite, its
null space is exactly the deformations that ``don't matter'' (rigid motion), and it
factors into rank-1 generalized plane quadrics whose ``normals'' are strain-sensing
co-modes. This is not a hack---it is the same template (preserve-a-quantity
$\rightarrow$ squared-linear-residual $\rightarrow$ sum) instantiated on strain instead
of distance.

\subsection{Where it genuinely diverges}

Two differences are not cosmetic; they are where the real design work lives.

\paragraph{1. The variable is a displacement (relative), not a position (absolute).}
Geometric fidelity is a \emph{zeroth-order} property of one point; mechanical fidelity is
a \emph{first-order} property---strain is a gradient, a \emph{difference} of
displacements. A single vertex's displacement determines no strain by itself. This is why
geometric QEM can collapse a vertex's incident faces into one portable $3\times3$ form on
that vertex's coordinate, but mechanics cannot: each element's quadric lives on its whole
stencil, and adjacent quadrics couple through shared DOFs. The additive structure still
exists---it is finite-element assembly---but it lives on the \emph{full} displacement
vector and never reduces to a per-vertex form. That coupling \emph{is} the mechanics; any
reduction that discards it discards what we are trying to preserve.

\paragraph{2. The face quadric is a \emph{response template}, not yet an \emph{error}.}
In geometric QEM the same quadric does double duty: it describes the plane \emph{and},
evaluated at the moved point, yields the error---because the original surface has zero
distance to its own planes, so any post-collapse distance \emph{is} the error.
Mechanically, $K_e$ describes the response, but the simplification error is the
\textbf{discrepancy} between fine and coarse response, which is not zero-referenced.
Concretely, with a prolongation $S$ mapping coarse displacements to fine ones, the
mechanically-exact coarse operator is the Galerkin form
\begin{equation}
  K_c = S^T K_f\,S,
\end{equation}
i.e.\ the fine-energy Hessian $K_f$ \emph{restricted} to the coarse displacement space
(a congruence reduction $S^T(\cdot)S$). The collapse error is how far the actual coarse
$K_c$ deviates from it---measured \emph{on some set of deformations}. The probe quadric
$G=V^TKV$ below is the same move with a different map $V$: restrict the Hessian to a small,
fixed probe space instead of to the coarse mesh.

That last phrase is the real fork: \textbf{which deformations do we probe the discrepancy
on?} This is the mechanical counterpart of ``which planes matter,'' and it is where
``responsiveness'' gets \emph{defined}:
\begin{itemize}
  \item \textbf{affine / linear displacement modes} $\rightarrow$ preserves the effective
        (homogenized) stiffness;
  \item \textbf{low-frequency eigenmodes of $K$} $\rightarrow$ preserves global compliance
        / vibration response;
  \item \textbf{a known load subspace} $\rightarrow$ preserves response to the loads we
        actually care about.
\end{itemize}
Choosing that probe subspace is what eventually lets us reduce the coupled assembly back
down to a small, portable, QEM-shaped quadric---i.e.\ it is the principled route to
additivity, rather than a hack imposed up front.
The remaining subsections develop this
program: we first establish \emph{why} the relative-displacement quadric resists
per-vertex reduction, then identify the two tools that buy reducibility back---static
condensation (Schur complement) for exact static fidelity, and a probe subspace for true
QEM portability---and finally show how the probe choice dials the mechanical quadric
between geometric QEM and an attribute-quadric.

\subsection{Why the mechanical quadric resists per-vertex reduction}

Geometric QEM reduces to a portable $4\times4$ form per vertex for two reasons, and
mechanics violates both.
\begin{itemize}
  \item \textbf{The residual is linear in a \emph{single} vertex's coordinate.}
        $r_f(\mathbf{x}) = \mathbf{n}^T\mathbf{x}+d$ involves one point. Strain
        $\boldsymbol\varepsilon = B\mathbf{d}$ is a discrete gradient---a \emph{difference}
        $\sim(\mathbf{u}_j-\mathbf{u}_i)/\ell$. A single node's displacement determines no
        strain, so the smallest object on which the residual is even defined is the element
        stencil, not a vertex.
  \item \textbf{The coefficients are frozen.} The plane $(\mathbf{n},d)$ is a constant
        reference baked in once at preprocessing. Mechanics has no frozen target
        displacement: the rest state stores zero energy, but so does \emph{every} rigid
        motion of it. The ``correct set'' is a whole null space (the rigid-body modes), not
        a point.
\end{itemize}
So the honest answer is that the mechanical quadric \textbf{does not reduce to a per-vertex
form in general}---this is exactly divergence~1 above. The quadric is irreducibly the
element form $K_e\in\mathbb{R}^{9\times9}$, and ``accumulation'' is finite-element assembly
$K=\sum_e K_e$ on the \emph{global} DOF vector. Reducibility can only be bought back in one
of two ways:
\begin{itemize}
  \item \textbf{Eliminate the vanishing DOF exactly} (Schur complement, \S\,below): keeps
        the full static response on the survivors; pays in fill-in.
  \item \textbf{Restrict to a frozen, per-vertex-factorable reference} (probe subspace,
        \S\,below): recovers a literal per-vertex additive quadric; pays in off-probe
        response.
\end{itemize}

\subsection{Schur complement: the mechanically-exact collapse operator}

An edge collapse \emph{removes} degrees of freedom, and static condensation is the exact
way to remove DOFs while preserving the response on what remains. Partition the global DOFs
into \emph{retained} ($r$) and \emph{eliminated} ($e$, the disappearing vertex's three
DOFs),
\begin{equation}
  K = \begin{pmatrix} K_{rr} & K_{re} \\ K_{er} & K_{ee} \end{pmatrix}.
\end{equation}
With no load on the eliminated DOFs they equilibrate to
$\mathbf{d}_e = -K_{ee}^{-1}K_{er}\mathbf{d}_r$, and the energy becomes
\begin{equation}
  \tfrac12\mathbf{d}^T K\mathbf{d}
    = \tfrac12\,\mathbf{d}_r^T\,\widehat{K}_{rr}\,\mathbf{d}_r,
  \qquad
  \widehat{K}_{rr} = K_{rr} - K_{re}K_{ee}^{-1}K_{er},
\end{equation}
where $\widehat{K}_{rr}$ is the \textbf{Schur complement}. Three facts make this the right
object:
\begin{itemize}
  \item \textbf{It is the Galerkin operator with the \emph{optimal} prolongation.}
        $\widehat{K}_{rr} = S^T K_f S$ when $S$ is the \emph{harmonic extension}
        (eliminated DOFs interpolated to minimize energy). So the Schur complement is not an
        alternative to $K_c = S^T K_f S$; it is the best-case $K_c$. Any actual collapse
        uses a simpler prolongation, and the collapse error is precisely
        $\|S_{\text{actual}}^T K_f S_{\text{actual}} - \widehat{K}_{rr}\|$.
  \item \textbf{It is local.} $K_{er}$ is nonzero only on the $1$-ring (FEM sparsity), so
        $\widehat{K}_{rr}$ differs from $K_{rr}$ only inside the $1$-ring. Eliminating a
        vertex is a $1$-ring operation---the locality QEM requires.
  \item \textbf{The error has a name: fill-in.} The correction
        $-K_{re}K_{ee}^{-1}K_{er}$ is a dense block coupling the neighbors of the
        eliminated vertex---Gaussian-elimination fill-in. Structurally, ``mechanical QEM''
        is approximate sparse Cholesky / algebraic-multigrid coarsening, and the quadric
        error of a collapse is the magnitude of the fill-in one refuses to store.
\end{itemize}

\subsection{Edge collapse as common-mode retention}

The collapse $(v_i,v_j)\to v^\star$ leaves the \emph{link} (the boundary of the edge's
$1$-ring) unchanged: same outer vertices, same outer edges. Only the interior is re-wired.
Mechanically, the boundary DOFs are fixed shared data, so the collapse is a local problem
with the link as a \textbf{Dirichlet boundary} and the merged vertex as the only free
interior DOF---exactly what the Schur complement needs.

To make ``the two vertices wiggle as one'' precise, change basis on the pair into a
\textbf{common mode} and a \textbf{differential mode},
\begin{align*}
  \mathbf{d}_+ &= \tfrac12(\mathbf{d}_i + \mathbf{d}_j)
    \quad(\text{move together}),
  \\
  \mathbf{d}_- &= \mathbf{d}_i - \mathbf{d}_j
    \quad(\text{stretch/shear the edge}).
\end{align*}
Two free vertices ($6$ DOF) split into $3$ common $+$ $3$ differential. The collapse
\textbf{keeps the common mode} ($\mathbf{d}^\star\leftrightarrow\mathbf{d}_+$) and
\textbf{removes the differential mode}---this is the $6\to3$ reduction. Hence the faithful
condition:
\begin{quote}
  The collapse is faithful iff, on the probe deformations, the differential mode
  $\mathbf{d}_-$ is unexcited---i.e.\ $v_i$ and $v_j$ were already moving together. The
  collapse cost is the elastic energy that lived in $\mathbf{d}_-$.
\end{quote}
Under a smooth probe field, adjacent vertices displace almost equally, so
$\mathbf{d}_-\approx 0$ and the collapse is cheap; this is why short edges in smooth regions
coarsen safely. It is also exactly algebraic-multigrid aggregation: lump together DOFs whose
relative motion is unexcited.

There are two distinct ways to remove $\mathbf{d}_-$, and the gap between them \emph{is} the
error:
\begin{itemize}
  \item \textbf{Restriction (what the collapse literally does):} set $\mathbf{d}_-=0$ and
        duplicate, $\mathbf{d}_i=\mathbf{d}_j=\mathbf{d}^\star$. Freezing a DOF can only
        \emph{add} stiffness, so a collapsed edge is always at least as stiff as the
        original on any retained motion.
  \item \textbf{Condensation (the ideal):} let $\mathbf{d}_-$ relax to its
        energy-minimizing value given the boundary $\mathbf{d}_b$ and common mode,
        \begin{equation}
          \mathbf{d}_-^\star
            = -K_{--}^{-1}\big(K_{-+}\mathbf{d}_+ + K_{-b}\mathbf{d}_b\big),
        \end{equation}
        which preserves the static response exactly (the Schur complement of the
        differential block).
\end{itemize}
The error is the gap between $\mathbf{d}_-=0$ and $\mathbf{d}_-=\mathbf{d}_-^\star$---the
energy recoverable by relaxing rather than freezing the edge. The merged position
$\mathbf{x}^\star$ (and the prolongation) is the one knob: choose it so the single vertex's
wiggle reproduces what the pair actually did, i.e.\ so that $\mathbf{d}_-^\star$ is small on
the probe. That \emph{is} the QEM minimization, restated.

\paragraph{Caveat.} ``Boundary same, interior re-wired'' is right for connectivity, but
three energy terms change, and only the first is the differential mode: (i) the relative
stretch/shear stored across the ring (the $\mathbf{d}_-$ energy---the main term); (ii) the
two collapsing triangles' own membrane and hinge energy (they go to zero area and vanish);
(iii) the dihedral angles of every ring edge shift, so the bending references in the ring
move. The differential-mode picture captures (i); (ii)--(iii) are why bending wants to be
re-evaluated at the collapse step, not frozen.

\subsection{Stretching and bending: triangle quadrics versus edge quadrics}

The two energies separate cleanly by stencil:
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{0.26\linewidth}|p{0.16\linewidth}|p{0.22\linewidth}|p{0.16\linewidth}|}
\hline
\textbf{energy} & \textbf{simplex} & \textbf{stencil} & \textbf{rank} \\
\hline
stretching (CST membrane) & triangle & 3 verts / 9 DOF & $3$ \\
\hline
bending (dihedral hinge) & interior edge & 4 verts / 12 DOF & $1$ \\
\hline
\end{tabular}
\end{center}
\noindent So bending enters at the \textbf{edge step}, as its own element family---not at
the triangle step. Two ways to handle it:
\begin{itemize}
  \item \textbf{Assembly view (recommended).} Assemble
        $K = \sum_{\text{tri}} K_e^{\text{stretch}} + \sum_{\text{edge}} K_e^{\text{bend}}$.
        Both families write into the $1$-ring blocks of the global operator, and the
        collapse (Schur or probe-reduction) operates on the \emph{merged} local operator.
        There is no ``triangle vs.\ edge step'' to decide because assembly already unifies
        them.
  \item \textbf{Garland-style per-vertex accumulation (if wanted for scoring).} Stretching
        distributes to its $3$ vertices like standard QEM; bending distributes to the
        hinge's $4$ stencil vertices (or its $2$ endpoints). But bending must be
        \emph{re-evaluated} at the collapse step, not frozen: a triangle's plane is
        collapse-invariant, whereas collapsing one edge perturbs the dihedral angle of every
        edge in the $1$-ring. Bending is therefore more fragile than stretching, with a
        wider stencil and a moving reference.
\end{itemize}
A deeper point connects to the probe discussion: \textbf{bending is invisible to
constant-strain (affine) deformation.} A flat patch under uniform in-plane strain does not
bend; bending is a second-derivative (curvature) quantity. So whether bending enters the
cost at all depends on the probe subspace---preserving bending requires enriching the probe
with quadratic (constant-curvature) modes $\mathbf{u}(X)=\tfrac12 X^T H X + FX + \mathbf{c}$.

\subsection{Affine probes and the homogenization patch test}

Take the probe subspace to be the \textbf{affine displacement fields}
$\mathbf{u}(X) = FX + \mathbf{c}$ with $F\in\mathbb{R}^{3\times3}$ constant. A constant-$F$
field produces \emph{constant strain} everywhere; the symmetric part of $F$ is strain, the
skew part is infinitesimal rotation, and $\mathbf{c}$ is translation (the last two are
rigid, zero-energy). The energetic content is the six independent constant-strain modes,
and the energy stored under a uniform macroscopic strain $\bar{\boldsymbol\varepsilon}$
\emph{is the definition} of the effective modulus,
\begin{equation}
  \tfrac12\,\bar{\boldsymbol\varepsilon}^T C^{\text{eff}}\,\bar{\boldsymbol\varepsilon}
    \cdot \mathrm{Vol} = \text{(total stored energy)}.
\end{equation}
Concretely, let the columns of $A$ be the affine modes sampled at the nodes. Then
\begin{equation}
  \text{``preserve effective stiffness''}
  \Longleftrightarrow
  A^T K_c A = A^T K_f A \ \text{on the patch,}
\end{equation}
which is precisely the finite-element \textbf{patch test}. This is the weakest but most
robust target: a six-dimensional subspace guaranteeing the coarse mesh is neither
artificially stiffer nor softer in bulk. It is blind to bending (see above), which is why
the eigenmode and load-subspace probes of the dictionary are needed when bending or
task-specific compliance matters.

\subsection{Reduction to an additive per-vertex quadric}

The payoff: committing to an affine probe \emph{collapses the irreducible coupling}, because
an affine field \textbf{factors through each vertex's own position}---reintroducing the
frozen, single-vertex reference that \S\,(why the quadric resists reduction) said mechanics
lacked. With $\mathbf{a} = (\mathrm{vec}(F),\mathbf{c})\in\mathbb{R}^{12}$, the displacement
at node $n$ is
\begin{equation}
  \mathbf{d}_n = F X_n + \mathbf{c}
    = \underbrace{[\,X_n^T\otimes I_3 \mid I_3\,]}_{P(X_n)\,\in\,\mathbb{R}^{3\times12}}
      \,\mathbf{a},
\end{equation}
depending only on node $n$'s rest position. The per-element probe energy is therefore
\begin{equation}
  W_e(\mathbf{a}) = \tfrac12\,\mathbf{a}^T G_e\,\mathbf{a},
  \
  G_e = V_e^T K_e V_e \in \mathbb{R}^{12\times12},
  \
  V_e = \begin{bmatrix} P(X_{n_1}) \\ P(X_{n_2}) \\ P(X_{n_3}) \end{bmatrix}.
\end{equation}
Two objects are worth naming. \textbf{$V_e$ is the probe sampler}: stacking $P(X_n)$ over
the element's nodes, it turns the $12$ affine parameters $\mathbf{a}$ into the element's $9$
nodal displacements, $\mathbf{d}=V_e\mathbf{a}$---a fixed linear map read off the rest
geometry alone (``evaluate this affine field at my corners''). \textbf{$G_e=V_e^T K_e V_e$
is then the element Hessian $K_e$ restricted to the affine subspace}: substitute
$\mathbf{d}=V_e\mathbf{a}$ into $\tfrac12\mathbf{d}^TK_e\mathbf{d}$ and the energy becomes
$\tfrac12\mathbf{a}^TG_e\mathbf{a}$. So $G_e$ is just $K_e$ \emph{seen through the probe}---the
stiffness the element shows against uniform strains---compressed from the awkward $9\times9$
stencil form to a portable $12\times12$ that lives in the same coordinates for every element.
The global probe stiffness $G = \sum_e G_e$ is the assembled Hessian under the same
restriction: it \textbf{accumulates additively over elements}---a small, fixed-size
($12\times12$, effective rank $6$, its null space the rigid modes), positive semi-definite
object that \emph{is} the homogenized stiffness in probe coordinates. The collapse cost is
the deviation of $G$ from its frozen fine-mesh target,
\begin{equation}
  E_{\text{collapse}}(\mathbf{x}^\star)
    = \big\| G_{\text{fine}} - G_{\text{after}}(\mathbf{x}^\star) \big\|_W^2
    = \Big\| \sum_{e\,\in\,\text{affected}}
        \big( G_e^{\text{old}} - G_e^{\text{new}}(\mathbf{x}^\star) \big) \Big\|_W^2,
\end{equation}
which is local, additive, fixed-size, and minimized over the merged position
$\mathbf{x}^\star$---structurally identical to a QEM scoring step. This is the Li-style
reduction: where Li attaches a fixed-size additive quadric $[\mathbf{p},\mathbf{u}]$ to each
vertex, here the per-vertex ``attribute'' is the vertex's probe-response signature, and the
metric is the energy metric $D$ rather than the identity.

\paragraph{The unifying picture.} The probe subspace dials the mechanical quadric between
two known forms:
\begin{itemize}
  \item \textbf{probe $=$ affine modes} $\rightarrow$ the signature is determined by
        position $\rightarrow$ the quadric reduces to (essentially) geometric QEM reweighted
        by the constitutive metric $D$; preserving homogenized stiffness under affine modes
        is mostly preserving area/shape, which geometric QEM already does;
  \item \textbf{probe $=$ low-frequency eigenmodes} $\rightarrow$ the signature is the
        mode-shape value per vertex $\rightarrow$ exactly Li's attribute quadric, but with
        the elastic mode shape as attribute and energy as metric---Nakagawa \& Kanai's
        ``modal strain as saliency'' turned into an actual quadric.
\end{itemize}
The mechanical $K_e$-in-the-$D$-metric is the energy-metric generalization that contains
both as probe-subspace special cases, with the Schur complement as the exact,
full-space limit.

\paragraph{Two honest caveats.} (i) $G_e(\mathbf{x}^\star)$ is not \emph{exactly} quadratic
in $\mathbf{x}^\star$, because moving the rest vertex changes $B$, the area, and hence $K_e$
nonlinearly; either freeze the fine reference and accept an approximate quadratic, or
evaluate at candidate placements (midpoint/endpoints) and take the minimum, as many QEM
variants do. (ii) The bargain is identical to geometric QEM's: response is preserved
\emph{within the probe subspace only}---off-probe (higher-frequency, non-affine) response is
discarded, just as geometric QEM preserves flatness but not higher-order curvature.

\subsection{The proposed pipeline}

We can now assemble the pieces into an end-to-end, QEM-shaped decimation pipeline. The
one-line summary: \emph{it is QEM, but the face quadric is replaced by the element stiffness
projected into a probe subspace ($G_e = V_e^T K_e V_e$), accumulated additively per vertex;
an edge collapse is priced by how much homogenized response on the probe it destroys
---equivalently, the energy in the differential mode it freezes out---and the merged vertex
is placed to minimize that loss.}

\paragraph{Phase 0 --- assemble the mechanical model.} On the fine mesh, assemble the two
element families into the global operator,
\begin{equation}
  K_f = \underbrace{\sum_{\text{tri}} A t\,B^T D B}_{\text{stretching (rank 3)}}
      \;+\; \underbrace{\sum_{\text{edge}} \tfrac12\,c\,\mathbf{g}\mathbf{g}^T}_{\text{bending hinge (rank 1)}}.
\end{equation}
This is the assembly view: both families write into the $1$-ring blocks; no
``triangle-vs-edge step'' decision is needed.

\paragraph{Phase 1 --- choose the probe subspace $V$.} This is where ``responsiveness'' is
\emph{defined}.
\begin{itemize}
  \item \textbf{Default:} the six affine / constant-strain modes $\rightarrow$ preserves the
        homogenized membrane stiffness (the patch test).
  \item \textbf{If bending matters:} enrich with the constant-curvature (quadratic) modes
        $\mathbf{u}(X)=\tfrac12 X^T H X + FX + \mathbf{c}$---affine probes are blind to bending.
  \item \textbf{If global compliance / known loads matter:} append low-frequency eigenmodes
        of $K_f$ or the load subspace.
\end{itemize}

\paragraph{Phase 2 --- build the mechanics-aware quadric (precompute).} Per element, project
the stiffness into probe coordinates, $G_e = V_e^T K_e V_e$, with $V_e$ sampling the probe
modes at the element's nodes through each node's rest position $P(X_n)$. Accumulate
additively to vertices, $G_v = \sum_{e \ni v} G_e$---the direct analog of QEM face-quadric
accumulation---and freeze the fine-mesh target $G_{\text{fine}}$.

\paragraph{Phase 3 --- greedy edge-collapse decimation.} Run the standard QEM heap loop. For
each candidate edge $(v_i, v_j)$:
\begin{enumerate}
  \item \textbf{Place} the merged vertex $\mathbf{x}^\star$ (and its probe signature) to
        minimize the mechanical cost---choosing it so the single vertex reproduces what the
        pair did on the probe, i.e.\ so the eliminated differential mode $\mathbf{d}_-$ stays
        small.
  \item \textbf{Score} with the composite cost
        \begin{equation}
          C \;=\; \underbrace{\big\|G_{\text{after}}(\mathbf{x}^\star) - G_{\text{fine}}\big\|_W^2}_{E_{\text{mech}}\ \text{(probe response lost)}}
            \;+\; \gamma\,E_{\text{triq}} \;+\; \alpha\,E_{\text{geom}},
        \end{equation}
        where $E_{\text{triq}}$ is the sliver penalty (required for a \emph{simulatable}
        output, per Nakagawa \& Kanai's ablation) and $E_{\text{geom}}$ is optional standard
        geometric QEM for visual fidelity.
  \item \textbf{Collapse} the heap minimum and \textbf{re-evaluate bending in the whole
        $1$-ring} (dihedral references move---bending is not collapse-invariant).
\end{enumerate}
Repeat to the target element budget. The merged vertex carries a position $\mathbf{x}^\star$
\emph{plus} a probe-response attribute---exactly Li's augmented vertex $[\mathbf{p},\mathbf{u}]$,
except the attribute is the elastic probe signature and the metric is $D$ rather than the
identity.

\paragraph{Forks still open.} The pipeline has a default at each branch, but the following
choices are genuine design/experiment forks:
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{0.24\linewidth}|p{0.42\linewidth}|p{0.24\linewidth}|}
\hline
\textbf{Fork} & \textbf{Options} & \textbf{Lean} \\
\hline
cost backbone & portable additive $\|G_{\text{after}}-G_{\text{fine}}\|$ vs.\ exact
  differential-mode Schur per collapse & (a); validate against (b) \\
\hline
probe subspace & affine / $+$ curvature / $+$ eigenmodes / $+$ loads & affine first \\
\hline
bending bookkeeping & fold into assembled $G_e$ vs.\ distribute to hinge's $4$ verts & assembled \\
\hline
merged placement & closed-form min of $G$-deviation vs.\ candidate set (mid/endpoints) & candidate set \\
\hline
metric $W$ & identity / energy-weighted / mode-importance & identity \\
\hline
\end{tabular}
\end{center}
Two checks de-risk the whole idea: (1) on a flat patch, the affine-probe additive quadric
should reduce to roughly $D$-reweighted geometric QEM (sanity); (2) on a single $1$-ring,
the additive cost should track the exact differential-mode Schur cost across collapse
candidates (correctness).

\subsection{Dirty meshes and simplicial complexes}

Real assets---a dress with dangling straps, lace, beads, non-manifold seams, disconnected
trim---stress the pipeline at its foundation: assembly. The load-bearing fact is that
\textbf{finite-element assembly is purely combinatorial and additive}: $K = \sum_e K_e$
scatters each element's dense block into the global matrix indexed by that element's DOFs,
and never asks whether an edge is manifold or a surface is closed. Everything below is a
consequence of that, and of where the consequence frays.

\paragraph{Constructing $K$ on a dirty mesh.} Assembly always succeeds; the real content is
the rule mapping each simplex to an element family, plus a convention at non-manifold edges.
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{0.30\linewidth}|p{0.30\linewidth}|p{0.30\linewidth}|}
\hline
\textbf{cell} & \textbf{element} & \textbf{quadric} \\
\hline
$0$-cell (bead) & point mass & --- (mass only) \\
\hline
$1$-cell (strap, lace) & bar / rod & rank-$1$ stretch ($+$ rod bending) \\
\hline
$2$-cell (panel) & membrane & rank-$3$ stretch; $+$ hinge on its edges \\
\hline
$3$-cell (solid) & P1 tetrahedron & rank-$6$ solid strain \\
\hline
\end{tabular}
\end{center}
Membrane stiffness assembles per triangle no matter how triangles are connected (a
non-manifold edge shared by $k$ panels just sums $k$ contributions into the shared DOFs). The
bending hinge is where non-manifoldness bites: a boundary edge ($1$ triangle) has \emph{no}
hinge (correct---a free hem has no bending partner), while a non-manifold edge ($k>2$
triangles) has an \textbf{ambiguous} dihedral angle and needs an explicit convention (skip /
sum all $\binom{k}{2}$ pairs / dominant pair). Dangling lower-dimensional simplices carry no
membrane and must be given bar/point elements if they are to carry load.

\paragraph{Is the operator ``proper''?} ``Proper'' splits into four senses that come apart on
dirty meshes.
\begin{itemize}
  \item \textbf{Symmetry $+$ PSD --- always holds.} Each $K_e$ is a $\tfrac12\|M\mathbf{d}\|^2$
        form, so $K=\sum K_e$ is symmetric PSD regardless of mesh quality. It is always a
        legitimate energy Hessian.
  \item \textbf{Null space --- grows, and the growth is diagnostic.} For a clean connected
        shell $\ker K$ is the six rigid modes. Each disconnected component adds six more;
        under-constraint adds \emph{spurious} zero modes (an isolated bead; a flat
        membrane patch, whose out-of-plane motion is zero-energy at linear order \emph{unless
        bending is present}; CST's missing drilling DOF; a bowtie vertex's counter-rotating
        cones). Hence bending is not optional for shells---it is what controls the
        membrane null space.
  \item \textbf{Conditioning --- dies on degeneracy.} Slivers give $K_e \sim 1/A \to \infty$:
        meaningless strain and a catastrophic condition number. This is why the triangle-quality
        term is load-bearing, not cosmetic.
  \item \textbf{Continuum consistency --- genuinely fails at non-manifold junctions.} There is
        no smooth $2$-manifold being discretized there, so ``convergence to the elasticity
        PDE'' is undefined. The operator remains a legitimate \emph{discrete} elastic energy,
        just not a Galerkin approximation of a shell.
\end{itemize}
The key reconciliation: our target $A^T K_c A = A^T K_f A$ references the fine
\emph{discrete} operator, so it stays well-defined even with no continuum limit. \textbf{The
mechanical-QEM objective degrades gracefully}---we never needed the continuum claim. The
fixes are exactly (i) a bending convention, (ii) null-space regularization, (iii) sliver
cleanup, and (iii) is already in the cost.

\paragraph{Collapse on simplicial complexes.} The three legs of the framework transfer
off-manifold. Assembly is combinatorial ($\checkmark$); the cost is linear algebra on $K$
($\checkmark$); and the affine probe still factors through vertex position at \emph{any}
cell's nodes, so $G_e = V_e^T K_e V_e$ accumulates additively across mixed-dimensional cells
---\textbf{the per-vertex additive quadric survives unchanged}. Mixed cells couple
automatically in assembly (a vertex shared by a panel and a strap sums both contributions),
which is just the ``fold bending and stretching into the assembled operator'' decision
generalized to bars and point masses. The topological scaffolding is prior art: Popovi\'c \&
Hoppe's \emph{Progressive Simplicial Complexes} (1997) generalizes vertex-split/edge-collapse
to arbitrary-dimension, arbitrary-topology complexes, and Garland \& Zhou's \emph{Quadric-Based
Simplification in Any Dimension} (2005) already puts the \emph{geometric} quadric on
$n$-simplices. The clean statement of our contribution: \textbf{inherit that contraction and
quadric machinery and swap the geometric quadric for the mechanical one.} Two caveats remain:
\begin{itemize}
  \item \textbf{Topological legality.} Simplicial contraction is governed by the
        \textbf{link condition}; violating it silently merges sheets, pinches, or kills
        handles. Controlled topology change (closing a sub-millimeter gap between panels) is
        often desirable, but must be a deliberate operator, not an accident.
  \item \textbf{Spurious null modes corrupt the cost.} A floppy/dangling region has a
        zero-energy mode, so collapsing it costs \emph{zero} mechanical error---``you cannot
        lose response you never had.'' The decimator will therefore aggressively delete exactly
        the dangling straps and flaps. This is why, on dirty assets, the geometric/feature
        term is \textbf{mandatory}: it protects mechanically-free but visually-important
        structure. The eigenmode probe inherits the same problem (the lowest modes of a dirty
        $K$ \emph{are} the floppy modes), which is a further reason to default to affine
        probes.
\end{itemize}

\paragraph{Volumetric decimation.} Extending from $2$-complexes (shells) to $3$-complexes
(solids) is, perhaps surprisingly, \emph{cleaner}. A P1 tetrahedron gives
$K_e = V\,B^T D B$ with the $6\times6$ solid constitutive $D$ and \textbf{no separate bending
family}---bending of a solid is just a strain distribution through its thickness, so the
awkward stretching/bending split vanishes: one element family, one quadric per tet. Solid
elasticity is a well-posed $3$D PDE with no non-manifold-junction ambiguity, so the patch test
and convergence are meaningful again, and interior-node elimination is textbook static
condensation (substructuring / superelements): \textbf{the exact Schur cost is on its firmest
footing volumetrically.} The operation gets stricter, though: tet edge-collapse must avoid
\textbf{inversion} (negative volume $\to$ indefinite $K_e$), so legality is the link condition
\emph{plus} a positive-volume constraint on all incident tets. And one usually wants to
preserve the boundary surface (visual) while coarsening the interior (mechanical)---a
two-criterion problem the simplicial-complex view handles natively: boundary triangles carry a
geometric quadric, interior tets carry the mechanical quadric, both accumulate to shared
vertices, the collapse cost sums them. The unifying structure is therefore a \textbf{single
abstract simplicial complex with a per-cell element type}, decimated by simplicial contraction
gated by (i) topological legality, (ii) geometric non-inversion, and (iii) the mechanical
cost. Surfaces and solids stop being separate pipelines.

\paragraph{The central hazard and a cheap diagnostic.} Across all of the above, the spurious
null space is the one thing we must \emph{design} rather than inherit. It is also cheaply
localizable: in $3$D,
\begin{equation}
  \big(\dim\ker K\big) - 6\,\big(\#\text{connected components}\big)
    \;=\; \#\,\text{spurious mechanism modes,}
\end{equation}
and visualizing \emph{where} those modes have support shows precisely the regions mechanics
would delete for free. Mitigations: a small regularization $K \to K + \epsilon M$ or a tiny
drilling/normal stiffness; always co-weighting a geometric/feature term; a cleanup pre-pass
(weld duplicates, drop zero-area/volume faces, snap near-coincident geometry) before any FEM.
The bottom line: the stiffness matrix can always be \emph{built} and is always a proper
\emph{discrete} energy; what a dirty mesh costs is continuum faithfulness (only at
non-manifold junctions, which our objective does not need) and conditioning (fixable by
cleanup).

\subsection{What the prototype computes, intuitively}

We implemented the affine/curvature instantiation of the program above as a working greedy
edge-collapse decimator. Stripped to intuition, it is six moves. The thread tying them
together: \textbf{the stiffness is the Hessian of the elastic energy}, and we coarsen the
mesh while preserving a probe-restricted version of that Hessian.

\paragraph{1. Two energies, one per simplex.} Each element stores energy only when the
deformation it ``cares about'' is excited; the element stiffness is the \textbf{Hessian} of
that energy.
\begin{itemize}
  \item \textbf{Stretching} lives on \emph{triangles}: the CST membrane
        $W_{\text{str}}=\tfrac12\,\mathbf{d}^\top K_e\,\mathbf{d}$ with
        $K_e=A\,t\,B^\top D B=\partial^2 W_{\text{str}}/\partial\mathbf{d}^2$ penalizes
        in-plane stretch/shear, weighted by area $A$, thickness $t$ and material $D$.
        Rank $3$, and \emph{blind} to out-of-plane motion.
  \item \textbf{Bending} lives on \emph{interior edges}: the hinge
        $W_{\text{bend}}=\tfrac12\,c\,(\mathbf{g}^\top\mathbf{d})^2$ with
        $\mathbf{g}=\nabla\theta$ penalizes change of the dihedral angle. Rank $1$; a
        boundary edge has no second triangle, hence no hinge.
\end{itemize}

\paragraph{2. The costs add because energies add.} Both families scatter into one global
elastic energy by finite-element assembly, $K=\sum_e K_e$---purely additive and
combinatorial, with no ``triangle step vs.\ edge step'' to decide. An edge collapse disturbs
only its $1$-ring, so every cost we compute is local.

\paragraph{3. A probe turns the response \emph{operator} into a \emph{number}.} $K$ is a
response operator, not an error: it vanishes at rest and on every rigid motion. To score a
collapse we must choose a family of test deformations---a \emph{probe}---and ask how the
response to them changes.
\begin{itemize}
  \item \textbf{Affine probe:} the uniform fields $\mathbf{u}(X)=FX+\mathbf{c}$. Their six
        constant-strain modes (symmetric $F$) carry energy; translation and rotation are
        rigid and free. Because an affine field reads each node only through its own rest
        position (via the sampler $V_e$, which maps the probe parameters to nodal
        displacements), the element response collapses to a small fixed-size quadric
        $G_e=V_e^\top K_e V_e$---\emph{the energy Hessian restricted to the probe}---that
        \textbf{accumulates additively per vertex}, exactly like a Garland--Heckbert quadric.
        Probing with these preserves the \emph{homogenized} (bulk) stiffness---the patch test.
  \item \textbf{Curvature probe:} an affine field is \emph{flat}---it sends a plane to a
        plane---so it leaves every dihedral angle untouched: a flat hinge's two triangles
        stay coplanar and the bending energy stored is \emph{exactly} zero. Affine probes are
        blind to bending for a structural reason: bending is a \emph{second} derivative of
        the displacement, and an affine field has none. So we enrich the probe with the
        smallest fields that carry a second derivative, the \textbf{constant-curvature
        (quadratic) fields} $u_k(X)=\tfrac12\,X^\top H_k X+F_{k,:}X+c_k$ (one symmetric
        Hessian $H_k$ per output component $k=1,2,3$), whose constant $H_k$ curves the patch
        the same way everywhere. This grows the probe from $12$ to $\mathbf{30}$ parameters:
        $6$ rigid $+$ $6$ constant-strain (the affine modes, still present) $+$ $\mathbf{18}$
        constant-curvature, and the eighteen new ones are what bend. The bookkeeping is
        \emph{unchanged}---$P(X)$ still reads each node through its own rest position, now
        with the quadratic monomials of $X$ appended (six per output component, $18$ in
        all)---so $G_e=V_e^\top K_e V_e$ is merely a larger $30\times30$ quadric that still
        \textbf{accumulates additively per vertex}; only the fixed size grows, $12\to30$.
        Why bending now registers: the hinge's quadric is
        $G_h=c\,(\mathbf{g}^\top V_e)^\top(\mathbf{g}^\top V_e)$, and its ``normal''
        $\mathbf{g}^\top V_e$ is the dihedral-angle change each probe mode causes. On a flat
        hinge an affine map keeps the triangles coplanar, so this vanishes on every affine
        mode---but a constant-curvature mode tilts the two triangles relative to each other,
        so it is \emph{nonzero} on the eighteen new modes. Bending therefore finally lives in
        $G$ and is priced at every collapse---which is exactly why a positive bending weight
        must be paired with the curvature probe. (It also tightens \emph{membrane} fidelity,
        since the in-plane quadratic modes pin strain \emph{gradients}, not just constant
        strain, at the cost of a stricter, ${\sim}6\times$ larger quadric.)
\end{itemize}

\paragraph{Probing in one tiny example (two springs).} Three nodes on a line, $A,m,B$ at
$x=0,1,2$, joined by two unit springs, so
$E=\tfrac12(u_m-u_A)^2+\tfrac12(u_B-u_m)^2$ and
\begin{equation}
  K=\begin{bmatrix}1&-1&0\\-1&2&-1\\0&-1&1\end{bmatrix}.
\end{equation}
The stiffness $K$ scores \emph{every} deformation, but it is $3\times3$ and entangled with
the interior node $m$. Probe it: the $1$-D affine field $\mathbf{u}(x)=\varepsilon x+c$
($\varepsilon$ a uniform stretch, $c$ a rigid translation), sampled at the nodes, gives the
probe map $V$ and the reduced stiffness $G$,
\begin{equation}
  V=\begin{bmatrix}x_A&1\\x_m&1\\x_B&1\end{bmatrix}
   =\begin{bmatrix}0&1\\1&1\\2&1\end{bmatrix},
  \qquad
  G=V^TKV=\begin{bmatrix}2&0\\0&0\end{bmatrix}.
\end{equation}
Read $G$ off directly: stretching the chain by a unit strain stores energy
$\tfrac12\cdot 2\cdot\varepsilon^2$, while translating it ($c$) stores nothing---the rigid
null mode. That single ``$2$'' is the chain's entire fingerprint as the probe sees it; the
$3\times3$ operator has been squeezed to one meaningful number (plus a structural zero).
\textbf{Collapse the interior node} and the coarse model is one spring $A\!-\!B$ of stiffness
$k_c$; probing it the \emph{same} way gives $G_c=k_c\,(x_B-x_A)^2=4k_c$---now directly
comparable, though the meshes differ in size. Matching the fingerprint forces
$k_c=\tfrac12$ (exactly the series stiffness $k_1k_2/(k_1{+}k_2)$), and the collapse cost
$\|G_c-G\|=|4k_c-2|$ rejects a naive reconnect ($k_c{=}1\Rightarrow$ cost $2$, too stiff) in
favor of the faithful one ($k_c{=}\tfrac12\Rightarrow$ cost $0$). The mesh version is
identical, only with the six constant-strain modes in place of the single $\varepsilon$, so
$G$ is $12\times12$ (six stiffness numbers $+$ six rigid zeros) rather than $2\times2$.

\paragraph{Choosing \emph{which} edge to collapse.} The score is not ``make it stiff''---it
is ``change the fingerprint as little as possible,'' and the two ways to remove $m$ are
generally \emph{not} equal. Give the springs different stiffnesses, $k_1=2$ on $A\!-\!m$ and
$k_2=1$ on $m\!-\!B$, so the fine fingerprint is $G_{\text{fine}}=k_1+k_2=3$. Removing $m$
collapses one of its two edges; the spring on the \emph{other} edge survives and now spans
the full $A\!-\!B$ gap (length $2$), so its probe energy uses that doubled span:
\begin{itemize}
  \item \textbf{collapse $A\!-\!m$} (the stiff edge): the soft spring survives,
        $G_{\text{after}}=k_2\,(x_B-x_A)^2=4$, cost $|4-3|=1$;
  \item \textbf{collapse $m\!-\!B$} (the soft edge): the stiff spring survives,
        $G_{\text{after}}=k_1\,(x_B-x_A)^2=8$, cost $|8-3|=5$.
\end{itemize}
The heap therefore collapses $A\!-\!m$ first---five times cheaper. (A single spring spanning
the whole gap is always stiffer than the two short ones it replaces, so both moves
over-stiffen; promoting the \emph{stiff} spring over-stiffens far more, so keeping the soft
one is the smaller change.) Note we are minimizing the \emph{deviation} of the fingerprint,
never maximizing stiffness: the score ranks the available moves and takes the gentlest. On a
mesh this is the same comparison, run over every edge with the $12$-number $G$.

\paragraph{Probing \emph{bending}: the same chain, now a beam.} The spring chain never
bends---axial springs store nothing in a sideways wiggle---so it cannot show what the
curvature probe is \emph{for}. Reuse the three nodes $A,m,B$ at $x=0,1,2$, but now let them
move \emph{transversely} (out of the line) by $w_A,w_m,w_B$, and add a discrete hinge that
penalizes the kink at the middle node,
\begin{equation}
  E_{\text{bend}}=\tfrac12\,c\,(w_A-2w_m+w_B)^2=\tfrac12\,c\,(\mathbf{g}^\top\mathbf{w})^2,
  \qquad \mathbf{g}=(1,-2,1)^\top .
\end{equation}
So $K_{\text{bend}}=c\,\mathbf{g}\mathbf{g}^\top$ is rank $1$---the $1$-D twin of the dihedral
hinge, with the discrete second difference $\mathbf{g}^\top\mathbf{w}$ standing in for
curvature. Probe it the \emph{affine} way first, with a transverse \emph{linear} field
$w(x)=\beta x+\gamma$ (a tilt $\beta$ plus a translation $\gamma$). A straight tilt has no
kink---its second difference is identically zero---so sampling at the nodes gives
$\mathbf{g}^\top V_{\!\text{aff}}=(0,0)$ and
\begin{equation}
  G_{\text{aff}}=V_{\!\text{aff}}^\top K_{\text{bend}} V_{\!\text{aff}}=\mathbf{0}.
\end{equation}
\textbf{The affine probe cannot see the hinge at all}---``blind to bending,'' here as a
literal $2\times2$ zero. Now enrich with the quadratic mode
$w(x)=\tfrac12\kappa x^2+\beta x+\gamma$; the sampler and the reduced quadric become
\begin{equation}
  V=\begin{bmatrix}\tfrac12 x_A^2 & x_A & 1\\ \tfrac12 x_m^2 & x_m & 1\\
                    \tfrac12 x_B^2 & x_B & 1\end{bmatrix}
   =\begin{bmatrix}0&0&1\\ \tfrac12&1&1\\ 2&2&1\end{bmatrix},
  \qquad
  G=V^\top K_{\text{bend}} V=\begin{bmatrix}c&0&0\\ 0&0&0\\ 0&0&0\end{bmatrix},
\end{equation}
in parameters $(\kappa,\beta,\gamma)$. The lone entry sits in the \emph{curvature} slot: the
second difference of the quadratic part is exactly $\kappa$ (so $\mathbf{g}^\top V=(1,0,0)$),
hence a unit curvature stores energy $\tfrac12 c$, while the tilt $\beta$ and translation
$\gamma$---the affine modes---fall in the null space and cost nothing. Exactly as the spring
chain squeezed its $K$ to the single stretch number ``$2$,'' the beam squeezes its hinge to
the single bending number ``$c$''---but only once the probe is rich enough to \emph{ask}
about curvature; the affine probe does not even carry that coordinate. (On a mesh this
$3\times3$ becomes the $30\times30$ $G$, with eighteen curvature modes in place of this lone
$\kappa$; and collapsing $m$ leaves the straight segment $A\!-\!B$ with no interior hinge, so
the coarse bending fingerprint is $0$ and the collapse forfeits the whole $c$---a curvature
response cannot be carried by fewer than three nodes, which is why bending guards features
that stretching ignores.)

\paragraph{4. The collapse cost is lost homogenized response.} Collapsing $(v_i,v_j)$ and
re-meshing the ring changes the accumulated $G$ by $\Delta G$; the price is
$\|\Delta G\|^2$---``how much homogenized response did this collapse destroy.'' Stretching,
bending and an optional geometric (QEM) term enter as \emph{separate} contributions, each
normalized to its own typical (median) magnitude so the weights are dimensionless---a weight
of $1$ means ``as costly as a typical membrane collapse.'' Bending is \emph{re-read at the
collapse}, not frozen, because the collapse shifts the dihedral references throughout the
ring.

\paragraph{5. A quiet fact: affine stretching $\approx$ geometry.} Preserving the homogenized
membrane stiffness under affine probes turns out to be essentially preserving area and
shape---i.e.\ geometric QEM reweighted by $D$. The cost penalizes pulling a vertex \emph{off}
the surface (quadratically) while letting it \emph{slide} tangentially (almost free), exactly
like a tangent-plane quadric---though we never wrote a distance-to-plane term. It
\emph{emerges} from patch-test invariance: re-triangulating a flat region in-plane does not
change its bulk stiffness, but tilting it out of plane does. So for membrane, mechanical and
geometric fidelity coincide; they part ways only where the material varies, or once the
curvature / eigenmode probes bring in bending and global compliance.

\paragraph{Corollary: boundaries for free.} The same area-sensitivity preserves \emph{free
boundaries} with no separate term. ``Slide tangentially for free'' holds in the interior
because the fan around a vertex is closed---an in-plane nudge shrinks some triangles and grows
others, leaving area (hence $G$) unchanged. At a boundary the fan is one-sided: sliding
\emph{along} the edge stays neutral, but sliding \emph{inward}---the in-surface direction
$\mathbf e\times\mathbf n$ perpendicular to the edge---can only shrink the boundary triangles,
with nothing on the open side to compensate, so it moves $G$ and is penalized. That penalized
direction is exactly the plane geometric QEM must \emph{add by hand} to stop boundary erosion
(its boundary-quadric normal is $\mathbf e\times\mathbf n$); here it falls out of preserving
stiffness, because a free edge is simply material that is one-sided. (Purely a membrane
effect---a boundary edge carries no hinge; and a \emph{dangling} flap is still a zero-energy
mode that only the geometric term guards.)

\paragraph{6. Where the merged vertex goes.} We keep it \emph{on the collapsing edge},
$w(\alpha)=(1-\alpha)\,\mathbf{x}_i+\alpha\,\mathbf{x}_j$ with $\alpha\in[0,1]$. The cost is
\emph{not} quadratic in a free $3$-D position (area and the strain gradient depend on it
nonlinearly), so there is no closed-form $-A^{-1}\mathbf{b}$ minimizer as in geometric QEM.
Instead we run a one-dimensional line search along the edge: evaluate the cost at
$\alpha\in\{0,\tfrac12,1\}$, fit a parabola, and take its minimum on $[0,1]$. Staying on the
edge keeps the coarse mesh on the original surface, and a placement that would sliver a
triangle prices itself out automatically (its area, hence its $G_e$, collapses and
$\|\Delta G\|$ jumps).

\medskip
\noindent The loop itself is ordinary greedy QEM---with the geometric face quadric swapped
for this mechanical, probe-projected $G$. It is worth spelling out \emph{when} and
\emph{where} each energy is evaluated:

\begin{quote}\small
\textbf{Precompute} (once, on the fine mesh) --- two element families: \\
\hspace*{1.5em} \emph{stretching}, on each \textbf{triangle} $f$ --- 3 verts, 9 DOF: \\
\hspace*{3em} $K_f=A\,t\,B^\top D B$ (rank 3); \emph{cache} $G_f=V_f^\top K_f V_f$ \\
\hspace*{1.5em} \emph{bending}, on each \textbf{interior edge} (hinge) $h$ --- 4 verts, 12 DOF: \\
\hspace*{3em} the 2 edge endpoints $+$ the 2 opposite triangle vertices, \\
\hspace*{3em} $K_h=c\,\mathbf{g}\mathbf{g}^\top$ (rank 1, $\mathbf{g}=\nabla\theta$); \emph{not} cached --- reassembled \\
\hspace*{3em} over the ring when an edge is scored (its dihedral reference drifts \\
\hspace*{3em} under every nearby collapse, so a per-hinge cache would be stale at once) \\
\hspace*{1.5em} normalizers $s_{\text{mem}},s_{\text{bend}},s_{\text{geom}}\leftarrow$ median initial per-edge term \\[3pt]
\textbf{Heap}: for every edge $(u,v)$, push $\textsc{cost}(u,v)$ \\[3pt]
\textbf{while} \#verts $>$ target: \\
\hspace*{1.5em} pop the cheapest edge $(u,v)$ and its stored $x^\star$;\ \ \textbf{skip} if stale or non-collapsible \\
\hspace*{1.5em} collapse $u,v\to x^\star$ \\
\hspace*{1.5em} \emph{refresh} the cache: recompute $G_f$ for the merged ring's triangles (geometry moved) \\
\hspace*{1.5em} re-push \emph{every} edge of that ring \hfill // not just edges at $x^\star$ \\[3pt]
\textbf{function} \textsc{cost}$(u,v)$: \hfill // local: only the ring changes; the rest cancels \\
\hspace*{1.5em} $A\leftarrow$ triangles incident to $u$ or $v$ \\
\hspace*{1.5em} $G^{\mathrm m}_{\text{old}}\leftarrow\sum_{f\in A}G_f$ \hfill // membrane, from the cache \\
\hspace*{1.5em} $G^{\mathrm b}_{\text{old}}\leftarrow$ assemble the ring's hinges at the current geometry \hfill // bending, fresh \\
\hspace*{1.5em} \textbf{for} each on-edge placement $x=(1-\alpha)\mathbf{x}_u+\alpha\mathbf{x}_v$, $\alpha\in\{0,\tfrac12,1\}$: \\
\hspace*{3em} $G^{\mathrm m}_{\text{new}}\leftarrow\sum$ \emph{surviving} triangles of $A$, each rebuilt with the vertex at $x$ \\
\hspace*{3em} $G^{\mathrm b}_{\text{new}}\leftarrow$ reassemble the ring's hinges with the merged geometry at $x$ \\
\hspace*{3em} $\text{mem}\leftarrow\|G^{\mathrm m}_{\text{new}}-G^{\mathrm m}_{\text{old}}\|^2$ \hfill // 2 collapsing triangles vanish \\
\hspace*{3em} $\text{bend}\leftarrow\|G^{\mathrm b}_{\text{new}}-G^{\mathrm b}_{\text{old}}\|^2$ \\
\hspace*{3em} $\text{geom}\leftarrow$ geometric-QEM error at $x$ \hfill // optional \\
\hspace*{3em} $C(\alpha)\leftarrow \text{mem}/s_{\text{mem}}+w_b\,\text{bend}/s_{\text{bend}}+w_g\,\text{geom}/s_{\text{geom}}$ \\
\hspace*{1.5em} fit a parabola to $C$ at $\alpha\in\{0,\tfrac12,1\}$ and also evaluate its minimizer $\alpha^\star\!\in[0,1]$ \\
\hspace*{1.5em} \textbf{return} the cheapest $C$ over $\alpha\in\{0,\tfrac12,1,\alpha^\star\}$, with its position $x$
\end{quote}

\noindent This is QEM's faces\,$\to$\,vertices\,$\to$\,edges flow, widened to two families.
\textbf{Stretching \emph{is} the QEM face quadric}---a triangle's $G_f$ scatters to its three
vertices---and it is the one piece cached. \textbf{Bending is a second family living on
interior edges}, rank-1, with a wider four-vertex stencil (the edge plus its two opposite
vertices); it is rebuilt locally rather than stored. And neither family is pre-accumulated into
a frozen per-vertex quadric the way classical QEM is: because the mechanical cost re-reads
geometry, the per-edge score sums the incident elements \emph{on the fly}. That is what keeps
the cost \emph{local}---every element outside the merged ring is identical before and after, so
it cancels in $G_{\text{new}}-G_{\text{old}}$ and is never touched. Of the membrane triangles
only the \emph{survivors} are rebuilt (the two folding onto the edge shrink to zero area and
drop out), and they are recomputed \emph{at the trial position $x$}, once per candidate
placement, since moving the vertex changes each $B$-matrix and area. Bending is likewise
reassembled at $x$---never frozen---because shifting one vertex tilts the dihedral reference of
every edge in the ring.

\medskip
\noindent The one caveat the pseudocode cannot hide: every energy above is evaluated
\emph{locally and now}. The placement line returns a faithful optimum for \emph{this} collapse
along \emph{this} edge, and the heap always spends the cheapest collapse available---but the
outer loop is greedy and path-dependent, so a locally cheaper collapse, or a locally better
$x^\star$, need not improve the \emph{global} coarse response: an early, gentle-looking
collapse can force expensive ones later. The per-step cost is an honest local score, not a
global optimum.

\subsection{Next steps: from the dihedral hinge to a plate-bending quadric}

The bending model is the least-developed part of the prototype, and a clean upgrade path falls
out of comparing it against \emph{Kirchhoff--Love} (KL) thin-plate theory. We record it here as
the next concrete experiment.

\paragraph{Where we are.} Membrane uses the plane-stress CST; bending uses the discrete dihedral
hinge (Grinspun et al., \emph{Discrete Shells}), $K_h=c\,\mathbf g\mathbf g^\top$ with
$\mathbf g=\partial\theta/\partial\mathbf v$ over the four hinge vertices. We already borrow KL's
flexural rigidity for the modulus, $c\propto Et^3/12(1-\nu^2)$. But the hinge has three
limitations, all traceable to its being a \emph{single scalar per edge}:
\begin{itemize}
  \item \textbf{Rank-1, one direction.} The dihedral measures bending only in the direction
        \emph{across} the edge (the edge is the fold axis)---a single directional sample of the
        curvature tensor $\kappa$, with no Poisson coupling between principal curvatures.
  \item \textbf{Under-uses the curvature probe.} Our $30$-D probe already spans the full
        constant-curvature tensor (the $18$ quadratic modes), but a rank-1 element excites only a
        $1$-D slice of it per edge---a rich probe paired with an impoverished element.
  \item \textbf{Transcendental in position.} $\theta(\mathbf x)$ is an arctan of the node
        positions, so the collapse cost is \emph{not} quadratic in the merged position
        $\mathbf x^\star$---exactly why placement needs the $1$-D edge line search instead of a
        QEM closed-form solve.
\end{itemize}

\paragraph{What KL offers.} KL bending energy is quadratic in the full \emph{curvature tensor}
$\kappa_{\alpha\beta}=-w_{,\alpha\beta}$ (the Hessian of the deflection), through a
moment--curvature law $M=D_b\kappa$ that uses the \emph{same} plane-stress matrix as membrane,
scaled by $t^3/12$ instead of $t$, \emph{with} Poisson coupling and the anticlastic $(1-\nu)$
term. The catch is continuity: a second-derivative energy needs $C^1$ shape functions (a slope
kink stores infinite bending energy), which plain $C^0$ triangles lack---hence non-conforming
plate elements (the Discrete Kirchhoff Triangle, DKT) or the constant quadratic-bending operator
of Bergou et al.\ (a cotangent bilaplacian $L$). The hinge is popular precisely because it sidesteps
$C^1$: it puts the curvature on edges as a fold angle and needs only vertex positions.

\paragraph{The upgrade, and why it is attractive.} Replace the hinge with a \textbf{linearized KL
curvature operator}: a fixed, rest-precomputed curvature--displacement matrix $B_b$ (DKT) or
constant $L$ (quadratic bending), so that $\kappa=B_b\,\mathbf x$ is \emph{linear} in the vertex
positions. Two payoffs follow at once.
\begin{itemize}
  \item \textbf{A rank-3, Poisson-coupled bending fingerprint.} $G_b=V^\top\!\big(B_b^\top D_b B_b
        \big)V$ becomes rank $3$ per element and carries the directional / Poisson structure the
        hinge cannot---a genuine \emph{bending patch test}, the clean analog of the membrane
        homogenization, and it finally exercises the full curvature probe. It also unifies the
        constitutive law (one $D$ for stretch and bend), opening the door to orthotropic materials
        the scalar hinge cannot express.
  \item \textbf{A QEM-shaped quadric in position.} Because $\kappa$ is linear in $\mathbf x$, the
        bending energy is a \emph{fixed quadratic form} $\tfrac12\,\mathbf x^\top Q_{\text{bend}}\,
        \mathbf x$ with $Q_{\text{bend}}=B_b^\top D_b B_b$, accumulating additively per vertex
        exactly like a Garland--Heckbert plane quadric (flat configurations are its null space).
        Combined with the membrane term---which already behaves like a tangent-plane quadric (the
        ``affine stretching $\approx$ geometry'' fact above)---the whole collapse cost becomes a
        single quadratic $C(\mathbf x^\star)=\mathbf x^{\star\top}\!A\,\mathbf x^\star+2\mathbf b^\top
        \mathbf x^\star+c$, with closed-form optimum $\mathbf x^\star=-A^{-1}\mathbf b$---recovering
        QEM's elegance and dropping the line search.
\end{itemize}

\paragraph{The bargain.} The fixed bending quadric is a \emph{linearization}: exact for a flat /
developable rest patch (where $\kappa=L\mathbf x$ is truly linear), approximate for curved rest or
large collapse moves, and not rotation-invariant under large deformation---the same trade QEM makes
by freezing its planes, and the same trade caveat~(i) already names (freeze the reference for an
approximate quadratic, or re-evaluate at candidates for accuracy). The hinge keeps rotation-invariance
and rest-exactness at the price of the line search; the linearized KL element buys closed-form
placement, a richer (rank-3, Poisson-coupled) operator, and a proper bending patch test at the price
of the linearization.

\paragraph{Concrete experiments.}
\begin{enumerate}
  \item Implement a per-triangle (or quadratic-bending) curvature operator $B_b$ and form
        $Q_{\text{bend}}=B_b^\top D_b B_b$; verify a \emph{bending patch test}---under a
        constant-curvature probe the homogenized $G_b$ matches the analytic plate stiffness, with
        the rigid and constant-strain modes in its null space.
  \item Project $B_b$ onto the curvature probe and confirm $G_b$ is rank $3$ (vs the hinge's rank
        $1$) on the crease demo; compare which features each preserves.
  \item Once $Q_{\text{bend}}$ is available, swap the bending line search for the closed-form
        $\mathbf x^\star=-A^{-1}\mathbf b$ and measure the fidelity / speed trade against the current
        re-evaluated, line-searched hinge cost.
\end{enumerate}
